Investigating the outer reaches of the Sculptor and Ursa Minor dwarf galaxies

### 2025-08-18 thesis/abstract.tex
thesis julio ch1 edits

Dwarf galaxies are unique cosmological probes, testing our understanding
of the smallest scales of structure formation in the Universe. Among the
classical dwarf galaxies, Sculptor and Ursa Minor appear unusual, with
more stars than expected in the outer regions compared with the typical
Exponential density profile, rememnescent of tidal interaction. I verify
the validity of this excess, checking \citet{jensen+2024}'s bayesian
membership catalogue for completeness, simple cuts, and other surveys.
To evaluate if tides indeed can explain these density features, I
conduct idealized N-body simulation of both galaxies in the tidal field
of the Milky Way. In both cases, the orbits are well constrained,
producing pericenters larger than expected to produce substantial tidal
impacts. In addition, the time since the last pericenter does not agree
with a tidal interpretation. However, Sculptor's orbit is strongly
influenced by the presence of an LMC. With an LMC, Sculptor's long term
orbital history is highly uncertain, however recent tidal effects due to
the Milky Way and LMC only minimally affect this galaxy. Because tides
likely do not explain the presence of these density features, I review
other theories about the formation of dwarf galaxy ``halos,'' including
mergers / accretion events and complex multi-component star formation
histories. The density profiles of Sculptor and Ursa Minor, with their
deviations from the classically expected exponential, present an
interesting puzzle, likely with a solution involving their formation and
history in the early universe.

>>>>

Dwarf galaxies are unique cosmological probes, which test our
understanding of the smallest scales of structure formation in the
Universe. \emph{Among the classical dwarf galaxies, we show that
Sculptor and Ursa Minor appear unusual, with more stars than expected in
the outer regions compared with the typical Exponential density
profile.}\footnote{need to refine this sentence} I verify the validity
of this excess, checking \citet{jensen+2024}'s bayesian membership
catalogue for completeness, simple cuts, and other surveys. To evaluate
if tides indeed can explain these density features, I conduct idealized
N-body simulation of both galaxies in the tidal field of the Milky Way.
In both cases, the orbits are well constrained, producing pericenters
larger than expected to produce substantial tidal impacts. In addition,
the time since the last pericenter does not agree with a tidal
interpretation. However, Sculptor's orbit is strongly influenced by the
presence of an LMC. With an LMC, Sculptor's long term orbital history is
highly uncertain, however recent tidal effects due to the Milky Way and
LMC only minimally affect this galaxy. Because tides likely do not
explain the presence of these density features, I review other theories
about the formation of dwarf galaxy ``halos,'' including mergers /
accretion events and complex multi-component star formation histories.
The density profiles of Sculptor and Ursa Minor, with their deviations
from the classically expected exponential, present an interesting
puzzle, likely with a solution involving their formation and history in
the early universe.

### 2025-08-18 thesis/data.tex
thesis julio ch1 edits

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{figures/scl_selection.pdf}}
\caption[Sculptor sample selection]{The selection criteria for Scl
members. Light grey points represent field stars (satisfying quality
criteria), turquoise points CMD+PM selected stars
(Eq.~\ref{eq:sel_cmd_pm}), blue xs probable members (2-component), and
RV confirmed members indigo diamonds. We mark the two stars from
\citet{sestito+2023a} with rust-outlined indigo stars. \textbf{Top:}
Tangent plane \(\xi, \eta\). The orange ellipse represents 3 half-light
radii. \textbf{Bottom left:} Colour magnitude diagram in Gaia \(G\)
magnitude versus \(G_{\rm BP} - G_{\rm RP}\) colour. We plot the Padova
12Gyr {[}Fe/H{]}=-1.68 isochone in orange. The black bar in the top left
represents the median colour error. \textbf{Bottom right:} Proper motion
in declination \(\mu_\delta\) vs RA \(\mu_{\alpha*}\) (corrected) the
orange circle represents the \citet{MV2020b} proper motion. The black
cross represents the median proper motion
error.}\label{fig:scl_selection}
\end{figure}

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{figures/umi_selection.pdf}}
\caption[Ursa Minor sample selection]{Similar to
Fig.~\ref{fig:scl_selection} except for Ursa Minor. We outline RV
members outside of \(6R_h\) in black stars (from \citet{sestito+2023b},
\citet{pace+2020} and \citet{spencer+2018}).}\label{fig:umi_selection}
\end{figure}

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{figures/fornax_selection.pdf}}
\caption[Fornax sample selection]{Similar to
Fig.~\ref{fig:scl_selection} except for Fornax. RV members are from
\citet{WMO2009}. While possibly a limitation of RV sample selection,
Fornax does not show the same extended outer halo of probable members as
Sculptor or Ursa Minor despite having many more
stars.}\label{fig:fornax_selection}
\end{figure}

>>>>

\begin{figure}
\centering
\includegraphics[width=0.7\linewidth,height=\textheight,keepaspectratio]{figures/scl_selection.png}
\caption[Sculptor sample selection]{The selection criteria for Scl
members. Light grey points represent field stars (satisfying quality
criteria), turquoise points CMD+PM selected stars
(Eq.~\ref{eq:sel_cmd_pm}), blue xs probable members (2-component), and
RV confirmed members indigo diamonds. We mark the two stars from
\citet{sestito+2023a} with rust-outlined indigo stars. \textbf{Top:}
Tangent plane \(\xi, \eta\). The orange ellipse represents 3 half-light
radii. \textbf{Bottom left:} Colour magnitude diagram in Gaia \(G\)
magnitude versus \(G_{\rm BP} - G_{\rm RP}\) colour. We plot the Padova
12Gyr {[}Fe/H{]}=-1.68 isochone in orange. The black bar in the top left
represents the median colour error. \textbf{Bottom right:} Proper motion
in declination \(\mu_\delta\) vs RA \(\mu_{\alpha*}\) (corrected) the
orange circle represents the \citet{MV2020b} proper motion. The black
cross represents the median proper motion
error.}\label{fig:scl_selection}
\end{figure}

\begin{figure}
\centering
\includegraphics[width=0.7\linewidth,height=\textheight,keepaspectratio]{figures/umi_selection.png}
\caption[Ursa Minor sample selection]{Similar to
Fig.~\ref{fig:scl_selection} except for Ursa Minor. We outline RV
members outside of \(6R_h\) in black stars (from \citet{sestito+2023b},
\citet{pace+2020} and \citet{spencer+2018}).}\label{fig:umi_selection}
\end{figure}

\begin{figure}
\centering
\includegraphics[width=0.7\linewidth,height=\textheight,keepaspectratio]{figures/fornax_selection.png}
\caption[Fornax sample selection]{Similar to
Fig.~\ref{fig:scl_selection} except for Fornax. RV members are from
\citet{WMO2009}. While possibly a limitation of RV sample selection,
Fornax does not show the same extended outer halo of probable members as
Sculptor or Ursa Minor despite having many more
stars.}\label{fig:fornax_selection}
\end{figure}

### 2025-08-18 thesis/discussion.tex
thesis julio ch1 edits

\citet{reed+gilmore2005} propsed that exponential dSph form through
impusive mass loss, redistributing the stellar component into an
exponential like profile.

@@

\citet{klimentowski2007}, \citet{klimentowski2009}.

@@

\begin{itemize}
\tightlist
\item
  \citet{arronyo-polonio+2024}
\item
  \citet{pace+2020}
\item
  \citet{fabrizio+2016}, \citet{kordopatis+2016} (multi-component in
  Carina)
\item
  \citet{battaglia+2006}, \citet{amorisco+evans2012} for fornax
\end{itemize}

@@

\begin{itemize}
\tightlist
\item
  \citet{deason+2014}
\item
  \citet{deason+2022}
\item
  \citet{ricotti+2022}
\item
  \citet{querci+2025}
\item
  \citet{amorisco+evans+vandeven2014}
\item
  \citet{tarumi+yoshida+frebel2021}: Shows cosmological simulations in
  context of Tuc II, merger of two galaxies in the early universe
  produces a more extended density profile and metallicity gradient.
\item
  \citet{lokas+2014}: And II from merger,
\item
  rotation from merger: \citet{cardona-barrero+2021}
\end{itemize}

>>>>

\citet{read+gilmore2005} propsed that exponential dSph form through
impusive mass loss, redistributing the stellar component into an
exponential like profile.

@@

\citet{klimentowski+2007}, \citet{klimentowski+2009}.

@@

\begin{itemize}
\tightlist
\item
  \citet{arroyo-polonio+2024}
\item
  \citet{pace+2020}
\item
  \citet{fabrizio+2016}, \citet{kordopatis+2016} (multi-component in
  Carina)
\item
  \citet{battaglia+2006}, \citet{amorisco+evans2012} for fornax
\end{itemize}

@@

\begin{itemize}
\tightlist
\item
  \citet{deason+2014}
\item
  \citet{deason+2022}
\item
  \citet{ricotti+polisensky+cleland2022}
\item
  \citet{querci+2025}
\item
  \citet{amorisco+evans+vandeven2014}
\item
  \citet{tarumi+yoshida+frebel2021}: Shows cosmological simulations in
  context of Tuc II, merger of two galaxies in the early universe
  produces a more extended density profile and metallicity gradient.
\item
  \citet{lokas+2014}: And II from merger,
\item
  rotation from merger: \citet{cardona-barrero+2021}
\end{itemize}
